```{r load data, include=FALSE}
variables <- readRDS("../data/load_variables.rds")
vars <- variables$data
list2env(vars, envir = .GlobalEnv)
```

# Findings {#sec-findings}

This chapter presents the findings from the structural topic modeling analysis of `r n_documents_after_aggregation` National Adaptation Plans. The analysis reveals both what countries discuss when planning for climate adaptation and how membership in different categories shapes these discussions. The patterns that emerge suggest systematic constraints on adaptation discourse that operate across diverse contexts.

The chapter has in two sections. The first examines the topics identified in the topic model and the thematic structure of global adaptation discourse. The second analyzes how category membership, by income, region, geography, and time, heavily influences topic prevalence, looking at both the concentration within the top topics, as well as STM's estimated effects.

Together, these findings reveal a paradox: despite removing the most common and rare words to focus on meaningful variation, adaptation discourse shows remarkable convergence around technical themes. This convergence varies systematically by category membership in ways that suggest institutional networks matter more than climate vulnerabilities in shaping how countries discuss adaptation.

## Topics

>This section examines the topics that structure global adaptation discourse. Each topic is characterized by its FREX terms, proportion of total discourse, and the countries where it features most prominently.

`r topic_8_name` commands an extraordinary `r topic_8_prop` of the entire corpus—more than triple any other topic—while appearing meaningfully in `r topic_8_documents` of `r n_documents_after_aggregation` national plans. The FREX terms "`r topic_8_frex`" reveal a discourse centered on planning infrastructure itself. The term _mainstream_ signals the integration of climate considerations into existing development frameworks, while _learn_ points to knowledge transfer mechanisms central to international climate architecture. References to the Paris Agreement (_pari_, _agreement_) situate this within global governance structures, and _progress_ likely refers to advancement metrics required by international reporting, other towards other goal.

The _Mainstream_ label captures how this topic represents the backbone of adaptation planning: the frameworks, agreements, and institutional arrangements that define what adaptation is, and can be. Its dominance suggests that regardless of whether countries face sea-level rise, desertification, or mountain glacial retreat, all must first engage with the technical governance language. The relationship between topic prevalence and document coverage reveals fundamental patterns in how adaptation discourse operates globally. @fig-topics maps each topic according to their share of total discourse (x-axis) and their presence across documents (y-axis), showing how the topics are distributed through the corpus.

```{r}
#| label: fig-topics
#| fig-cap: "Topics by proportion and document coverage. The mainstream topic dominates both dimensions."
#| fig-height: 5
#| echo: false

# Load dependencies
library(ggplot2)
library(ggrepel)
library(ggthemes)
library(scales)

# Load data
topics <- readRDS("../data/name_topics.rds")
topics_table <- topics$data

# Create plot
ggplot(topics_table, aes(x = topic_proportion, y = effective_documents)) +
  geom_point(size = 2) +
  geom_text_repel(
    aes(label = topic_name),
    force = 4,
    point.padding = 0.5,
    box.padding = 0.5,
    size = 5
  ) +
  labs(
    title = "Topic distribution",
    x = "Topic Proportion",
    y = "Effective Documents"
  ) +

  scale_x_continuous(labels = percent_format(accuracy = 1)) +

  theme_economist_white(base_size = 11) +
  theme(
    legend.position = "none",
    plot.title = element_text(
      size = 16,
      face = "bold",
      margin = margin(b = 15),
      hjust = 0.5
    ),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10))
  )
```

The remaining topics address diverse vulnerabilities and planning dimensions, though all filtered through institutional lenses and operating at much smaller scales. 

`r topic_1_name` occupies `r topic_1_prop` of discourse and appears meaningfully in `r topic_1_documents` documents. It has the FREX terms "`r topic_1_frex`" and features the most in `r topic_1_countries`. The fragmented nature of most terms gives some doubts about the validity of the topic or as artifacts from the word-splitting that might have been in the pdfs. _Ment_ (might be from development, management, government), _Tion_ (adaptation, mitigation, implementation), and _Pro_ (from project, program, process), alongside _Percent_ still suggests that this topic captures some planning language. Only _Napa_ provides clear thematic content, referencing the National Adaptation Programme of Action framework, the predecessor to the NAP [@mizuno2024].

`r topic_2_name` comprises `r topic_2_prop` of the corpus, present in `r topic_2_documents` documents, with FREX terms "`r topic_2_frex`" dominated by `r topic_2_countries`. The coherent terminology around tropical weather systems (_Cyclon_, _Tropic_) and temporal patterns (_Decad_, _Sea_, _Rise_) validates this topic's thematic integrity. The _Cyclone_ name reflects the focus on vulnerabilities central to many coastal and island nations.

`r topic_3_name` represents only `r topic_3_prop` of discourse across `r topic_3_documents` documents, with FREX terms "`r topic_3_frex`" and presence in `r topic_3_countries`. The terms mix governmental structures (_Republ_, _Feder_) with geography (_Mountain_), seasonality (_Summer_), and formal agreements (_Accord_). This captures federal republics discussing mountain environments and seasonal variations, and the _Mountain_ label is appropriate.

`r topic_4_name` at `r topic_4_prop` appearing in `r topic_4_documents` documents shows FREX terms "`r topic_4_frex`" concentrated in `r topic_4_countries`. The marine and coastal terminology (_Marin_, _Island_, _Hurrican_, _Coastal_, _Mangrov_) is a coherent set of terms around coastal vulnerabilities. The *Hurricane* name captures this.

`r topic_5_name` constitutes `r topic_5_prop` found in `r topic_5_documents` documents with FREX terms "`r topic_5_frex`" prominent in `r topic_5_countries`. The bureaucratic vocabulary (_Offic_, _Climate-Resili_, _Medium_, _Secretariat_, _Depart_) reveals institutional architecture considerations, with _Climate-Resili_ suggesting these offices focus specifically on resilience building. The _Office_ label captures this administrative focus.

`r topic_6_name` at `r topic_6_prop` in `r topic_6_documents` documents with FREX terms "`r topic_6_frex`" dominated by `r topic_6_countries`. The terminology spans infrastructure (_Transit_), planning processes (_Task_, _Instrument_), and indigenous rights (_Indigen_, _Territori_). The _Transit_ label captures only one dimension of what appears to be a complex intersection of infrastructure development and indigenous territorial governance. Brazil and Argentina's dominance suggests this topic reflects Latin American contexts where infrastructure projects navigate indigenous rights.

`r topic_7_name` comprises `r topic_7_prop` across `r topic_7_documents` documents with FREX terms "`r topic_7_frex`" appearing in `r topic_7_countries`. The technical climate science vocabulary (_Rcp_ for Representative Concentration Pathways, _Rainfal_, _Ensembl_, _Confid_, _Trend_) forms the clearest thematic cluster, representing the modeling and projection work underlying vulnerability assessments. The _RCP_ name identifies this as the technical scientific foundation of adaptation planning.

```{r}
#| label: tbl-topics
#| tbl-cap: "Summary of adaptation discourse topics identified through structural topic modeling"
#| tbl-colwidths: [15, 10, 10, 35, 30]
#| echo: false

# Load required packages
library(knitr)
library(kableExtra)

# Create topics data frame using a loop
topics_data <- data.frame()

for (i in 1:k_topics) {
  topics_data <- rbind(
    topics_data,
    data.frame(
      Topic = get(paste0("topic_", i, "_name")),
      Proportion = get(paste0("topic_", i, "_prop")),
      Documents = get(paste0("topic_", i, "_documents")),
      `Key Terms` = get(paste0("topic_", i, "_frex")),
      Countries = get(paste0("topic_", i, "_countries"))
    )
  )
}

# Create formatted table (already in topic number order from loop)
kable(
  topics_data,
  row.names = FALSE,
  col.names = c("Topic", "Prop", "Docs", "Terms", "Countries")
)
```

## Groups

>This section examines how category membership shapes adaptation discourse, showing that regional institutional networks have a much stronger influence than geography, income level, or time period.

Regional groupings demonstrate the most powerful influence on adaptation discourse, with an average effect size of `r region_effect`, much higher than income (`r income_effect`), geography (`r geography_effect`), or time (`r time_effect`). This dominance suggests that regional institutional architectures, like development banks, technical assistance programs, or consultant networks, create a convergence around climate knowledge that outweighs all the other factors.

Europe & Central Asia exhibit the strongest regional effect at `r europe_central_asia_effect`, nearly half the total possible effect size. Countries in this region show dominance of `r europe_central_asia_dom` concentrated on `r europe_central_asia_topics`. Despite the region being very large and diverse, from EU members to Central Asian republics, these plans are very similar. The emphasis on mountain ecosystems alongside mainstream planning suggests either genuinely shared vulnerabilities or adoption of standardized regional templates and approaches- 

South Asia follows with an effect size of `r south_asia_effect`, the second highest among regions. With only four plans in the dataset, this region nonetheless shows dominance of `r south_asia_dom` focused on `r south_asia_topics`. The concentration suggest that the nations have institutions or other shared regional resources in common.

Sub-Saharan Africa and Middle East, North Africa, Afghanistan & Pakistan both show effect sizes of `r sub_saharan_africa_effect`. Sub-Saharan Africa's dominance of `r sub_saharan_africa_dom` through `r sub_saharan_africa_topics` reflects the region's longer engagement with adaptation planning through NAPA frameworks. Meanwhile, MENA's lower dominance of `r middle_east_north_africa_afghanistan_pakistan_dom` but equal effect size suggests that while this region explores more diverse topics, countries within it still converge strongly around `r middle_east_north_africa_afghanistan_pakistan_topics`.

The relationship between discourse dominance and group effect sizes reveals how different categorization schemes create varying levels of constraint on national adaptation planning. @fig-dominance-effect maps these patterns across all subcategories.

```{r}
#| label: fig-dominance-effect
#| fig-title: "Dominance and effect"
#| fig-cap: "Dominance and Effect Size by Category. Each panel shows a different categorization scheme."
#| fig-height: 6
#| echo: false

# Load required packages
library(ggplot2)
library(ggrepel)
library(ggthemes)

# Load metrics data
metrics <- readRDS("../data/calculate_metrics.rds")
metrics_table <- metrics$data

# Prepare data for plotting
plot_data <- metrics_table[
  metrics_table$subcategory != "Overall" & !is.na(metrics_table$effect_size),
]
plot_data$category <- factor(
  plot_data$category,
  levels = c("Income", "Region", "Geography", "Time")
)
plot_data$label <- ifelse(plot_data$significant, plot_data$subcategory, "")

ggplot(plot_data, aes(x = dominance, y = effect_size)) +
  geom_point(size = 2) +
  geom_text_repel(
    aes(label = subcategory_abbrev),
    force = 3,
    point.padding = 0.5,
    box.padding = 0.5,
    size = 3
  ) +
  labs(
    title = "Dominance and effect",
    x = "Dominance",
    y = "Estimated effect"
  ) +
  facet_wrap(~category, scales = "fixed", ncol = 2) +

  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +

  theme_economist_white(base_size = 11) +
  theme(
    plot.title = element_text(
      size = 16,
      face = "bold",
      margin = margin(b = 15),
      hjust = 0.5
    ),
    strip.text = element_text(size = 12, face = "bold"), # Facet labels
    panel.spacing = unit(1, "lines"), # increases space between panels
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10))
  )
```

High-income countries show extreme topic concentration (`r high_income_dom`) but moderate group effects (`r high_income_effect`), while low-income countries display lower dominance (`r low_income_dom`) but the highest income-group effect (`r low_income_effect`). This suggests that discourse concentration and group conformity might not be directly related. Figure @fig-average presents the aggregate patterns, confirming regional categories as the primary organizing force in global adaptation discourse.

```{r}
#| label: fig-average
#| fig-cap: "Average dominance and effect size by category. Regional patterns show the strongest effects."
#| fig-height: 5
#| echo: false
# Load dependencies
library(tidyr)

# Filter and select
temp_data <- metrics_table[
  metrics_table$subcategory == "Overall" & metrics_table$category != "Global",
]
temp_data <- temp_data[, c("category", "dominance", "effect_size")]

# Reorder category by dominance
temp_data$category <- reorder(temp_data$category, temp_data$dominance)

# Pivot longer
plot_data <- pivot_longer(
  temp_data,
  cols = c(dominance, effect_size),
  names_to = "metric",
  values_to = "value"
)

# Factor the metric
plot_data$metric <- factor(
  plot_data$metric,
  levels = c("effect_size", "dominance")
)

# Create plot
ggplot(plot_data, aes(x = value, y = category, fill = metric)) +
  geom_col(position = "dodge", alpha = 0.7) +
  geom_text(
    aes(label = paste0(round(value * 100), "%")),
    position = position_dodge(width = 1),
    hjust = -0.1,
    size = 4
  ) +
  labs(title = "Average across categories", x = "Average Value", y = NULL) +
  scale_x_continuous(
    expand = expansion(mult = c(0, 0.1)),
    labels = percent_format(accuracy = 1)
  ) +

  theme_economist_white(base_size = 11) +
  scale_fill_economist(name = "", labels = c("Effect Size", "Dominance")) +
  theme(
    legend.position = c(0.98, 0.02), # Bottom-right corner
    legend.justification = c(1, 0), # Anchor to bottom-right
    legend.background = element_rect(fill = "transparent", color = NA),
    legend.title = element_blank(),
    legend.text = element_text(size = 10), # Match base_size
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    plot.title = element_text(
      size = 16,
      face = "bold",
      margin = margin(b = 15),
      hjust = 0.5
    )
  )
```

The consistently high values across both metrics for regions, compared to the variation in other categories, reinforces that institutional geography trumps physical geography in shaping how countries articulate climate responses.

Income categories show an overall effect size of `r income_effect`, revealing counterintuitive patterns. Low-income countries demonstrate the highest effect within this category at `r low_income_effect`—nearly double that of upper-middle income countries (`r upper_middle_income_effect`). This extreme convergence among poor countries, despite their geographic and climatic diversity, suggests powerful standardizing forces operate through development assistance and technical support mechanisms.

Low-income countries focus on `r low_income_topics` with dominance of `r low_income_dom`. The prominence of NAPA, the old NAP process, alongside mainstream planning might reflect how these countries have been adapting for a while, or that they must demonstrate fluency in both historical and contemporary adaptation frameworks. Their limited autonomy in discourse construction likely reflects dependence on external technical assistance and the need to align with donor priorities. High-income countries show dominance of `r high_income_dom` concentrated on `r high_income_topics`, yet their effect size of `r high_income_effect` sits mid-range. This combination of extreme topic concentration with moderate group coherence suggests that while wealthy countries converge on certain topics, they maintain some flexibility in how they articulate them. Upper-middle income countries have the room to maneuver in their plans, with the lowest dominance (`r upper_middle_income_dom`) and smallest effect size (`r upper_middle_income_effect`) focused on `r upper_middle_income_topics`. 

Geographic vulnerability classifications show an overall effect of `r geography_effect`, lower than the other categories. Countries sharing similar climate vulnerabilities show less discourse similarity than countries in the same World Bank region, challenging assumptions about environmental determinism in adaptation planning.

Small Island Developing States exhibit the highest dominance in the entire dataset at `r sids_dom`, focusing on `r sids_topics` with an effect size of `r sids_effect`. This concentration seems to reflect shared threats from sea-level rise and tropical cyclones. Yet even here, `r topic_8_name` dominates alongside obviously relevant island concerns, suggesting institutional requirements constrain even the most vulnerable nations. Landlocked Developing Countries show moderate patterns with dominance of `r lldc_dom` through `r lldc_topics` and effect size of `r lldc_effect`. Surprisingly, all of the topics are technical and governance-related, despite the group being defined by their geography. Other countries display the lowest geographic dominance at `r other_dom` with an effect size of `r other_effect` focused on `r other_topics`.

The time periods have the weakest effects at `r time_effect`, suggesting adaptation discourse has not substantially evolved over the decade of NAP development. While individual periods show variation—middle period plans (2019-2021) demonstrating highest dominance at `r middle_dom` focused on `r middle_topics`—the overall temporal effect remains minimal. 

Early submissions (2015-2018) show dominance of `r early_dom` with the lowest temporal effect of `r early_effect` on `r early_topics`. Middle period submissions exhibit the highest temporal convergence at `r middle_effect`, coinciding with clearer guidelines and more rigid finance requirements. Late submissions (2022-2025) display the lowest dominance at `r late_dom` but maintain substantial effect size of `r late_effect` on `r late_topics`. The temporal patterns reveal no clear trajectory toward either convergence or divergence, suggesting that once established, adaptation planning frameworks persist regardless of accumulating experience or evolving climate science. 

```{r}
#| label: tbl-groups
#| tbl-cap: "Discourse patterns by category membership showing dominance and effect sizes"
#| tbl-colwidths: [15, 30, 15, 10, 30]
#| echo: false

# Load required packages
library(knitr)
library(kableExtra)

# Define all subcategories and their parent categories
subcategories <- list(
  Income = c(
    "high_income",
    "upper_middle_income",
    "lower_middle_income",
    "low_income"
  ),
  Region = c(
    "europe_central_asia",
    "latin_america_caribbean",
    "south_asia",
    "sub_saharan_africa",
    "east_asia_pacific",
    "middle_east_north_africa_afghanistan_pakistan"
  ),
  Geography = c("sids", "lldc", "other"),
  Time = c("early", "middle", "late")
)

# Create clean name mapping
clean_names <- c(
  "high_income" = "High Income",
  "upper_middle_income" = "Upper-Middle Income",
  "lower_middle_income" = "Lower-Middle Income",
  "low_income" = "Low Income",
  "europe_central_asia" = "Europe & Central Asia",
  "latin_america_caribbean" = "Latin America & Caribbean",
  "south_asia" = "South Asia",
  "sub_saharan_africa" = "Sub-Saharan Africa",
  "east_asia_pacific" = "East Asia & Pacific",
  "middle_east_north_africa_afghanistan_pakistan" = "MENA, Afghanistan & Pakistan",
  "sids" = "Small Island Developing States",
  "lldc" = "Landlocked Developing Countries",
  "other" = "Other",
  "early" = "Early (2015-2018)",
  "middle" = "Middle (2019-2021)",
  "late" = "Late (2022-2025)"
)

# Create data frame with single loop
group_data <- data.frame()

for (cat in names(subcategories)) {
  for (subcat in subcategories[[cat]]) {
    group_data <- rbind(
      group_data,
      data.frame(
        Category = cat,
        Subcategory = clean_names[subcat],
        Dominance = get(paste0(subcat, "_dom")),
        `Effect Size` = get(paste0(subcat, "_effect")),
        `Top Topics` = get(paste0(subcat, "_topics"))
      )
    )
  }
}

# Create formatted table
kable(
  group_data,
  format = "markdown",
  row.names = FALSE,
  escape = FALSE,
  col.names = c("Category", "Group", "Dominance", "Effect", "Topics")
)
```
