```{r load data, include=FALSE}
variables <- readRDS("../data/load_variables.rds")
vars <- variables$data
list2env(vars, envir = .GlobalEnv)
```

# Findings {#sec-findings}

This chapter presents the findings from the structural topic modeling analysis of `r n_documents_after_aggregation` National Adaptation Plans. The analysis reveals both what countries discuss when planning for climate adaptation and how membership in different categories shapes these discussions. The patterns that emerge suggest systematic constraints on adaptation discourse that operate across diverse contexts.

The chapter has in two sections. The first examines the topics identified in the topic model and the thematic structure of global adaptation discourse. The second analyzes how category membership, by income, region, geography, and time, heavily influences topic prevalence, looking at both the concentration within the top topics, as well as STM's estimated effects.

Together, these findings reveal a paradox: despite removing the most common and rare words to focus on meaningful variation, adaptation discourse shows remarkable convergence around technical themes. This convergence varies systematically by category membership in ways that suggest institutional networks matter more than climate vulnerabilities in shaping how countries discuss adaptation.
## Topics

>This section examines the topics that structure global adaptation discourse. Each topic is characterized by its FREX terms, proportion of total discourse, and the countries where it features most prominently.

`r topic_8_name` commands an extraordinary `r topic_8_prop` of the entire corpus—more than triple any other topic—while appearing meaningfully in `r topic_8_documents` of `r n_documents_after_aggregation` national plans. The FREX terms "`r topic_8_frex`" reveal a discourse centered on planning infrastructure itself. The term _mainstream_ signals the integration of climate considerations into existing development frameworks, while _learn_ points to knowledge transfer mechanisms central to international climate architecture. References to the Paris Agreement (_pari_, _agreement_) situate this within global governance structures, and _progress_ likely refers to advancement metrics required by international reporting, other towards other goal.

The _Mainstream_ label captures how this topic represents the procedural backbone of adaptation planning: the frameworks, agreements, and institutional arrangements that define what adaptation can be. Its dominance suggests that regardless of whether countries face sea-level rise, desertification, or mountain glacial retreat, all must first demonstrate fluency in this technical governance language to access resources and legitimacy. The relationship between topic prevalence and document coverage reveals fundamental patterns in how adaptation discourse operates globally. @fig-topics maps each topic according to both their share of total discourse and their presence across documents, exposing how certain frameworks achieve hegemonic status while others remain peripheral.


```{r}
#| label: fig-topics
#| fig-cap: "Topics by proportion and document coverage. The mainstream topic dominates both dimensions."
#| fig-height: 5
#| echo: false

# Load dependencies
library(ggplot2)
library(ggrepel)
library(ggthemes)
library(scales)

# Load data
topics <- readRDS("../data/name_topics.rds")
topics_table <- topics$data

# Create plot
ggplot(topics_table, aes(x = topic_proportion, y = effective_documents)) +
  geom_point(size = 2) +
  geom_text_repel(aes(label = topic_name), 
            force = 4,
            point.padding = 0.5,
            box.padding = 0.5,
            size = 5) +
  labs(title = "Topic distribution",
       x = "Topic Proportion",
       y = "Effective Documents"
  ) +
  
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  
  theme_economist_white(base_size = 11) + 
  theme(legend.position = "none",
        plot.title = element_text(size = 16, face = "bold", 
                             margin = margin(b = 15), 
                             hjust = 0.5),
        axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 10)))
```

The remaining topics address diverse vulnerabilities and planning dimensions, though all filtered through institutional lenses and operating at much smaller scales. 

`r topic_1_name` occupies `r topic_1_prop` of discourse, appearing meaningfully in `r topic_1_documents` documents, with FREX terms "`r topic_1_frex`" and concentration in `r topic_1_countries`. The fragmented nature of most terms—_Ment_ (likely from development, management, government), _Tion_ (from adaptation, mitigation, implementation), and _Pro_ (from project, program, process)—alongside _Percent_ suggests this topic captures formulaic planning language. Only _Napa_ provides clear thematic content, referencing the National Adaptation Programme of Action framework. The _NAPA_ label is appropriate given this is the only complete meaningful term, though the topic appears to represent linguistic patterns common in countries with NAPA experience rather than coherent subject matter.

`r topic_2_name` comprises `r topic_2_prop` of the corpus, present in `r topic_2_documents` documents, with FREX terms "`r topic_2_frex`" dominated by `r topic_2_countries`. The coherent terminology around tropical weather systems (_Cyclon_, _Tropic_) and temporal patterns (_Decad_, _Sea_, _Rise_) validates this topic's thematic integrity. The _Cyclone_ designation accurately reflects the focus on tropical storm vulnerabilities central to many coastal and island nations. That such acute climate threats occupy less than 10% of global adaptation discourse reveals how even the most urgent vulnerabilities must compete for space within institutionally-defined frameworks.

`r topic_3_name` represents only `r topic_3_prop` of discourse across `r topic_3_documents` documents, with FREX terms "`r topic_3_frex`" and presence in `r topic_3_countries`. The terminology mixing governmental structures (_Republ_, _Feder_), geography (_Mountain_), seasonality (_Summer_), and formal agreements (_Accord_) suggests this captures federal republics discussing mountain environments and seasonal variations. The _Mountain_ label partially captures the geographic focus, though the prominence of governmental terms indicates this may reflect how certain political structures (federal republics) approach highland adaptation challenges.

`r topic_4_name` at `r topic_4_prop` appearing in `r topic_4_documents` documents shows FREX terms "`r topic_4_frex`" concentrated in `r topic_4_countries`. The marine and coastal terminology (_Marin_, _Island_, _Hurrican_, _Coastal_, _Mangrov_) forms a highly coherent cluster around island and coastal vulnerabilities. The _Island_ designation perfectly captures this topic's focus on maritime environments and the specific challenges of island states facing hurricanes and coastal degradation. Its relatively modest presence despite clear thematic coherence suggests even well-defined vulnerabilities receive limited discursive space.

`r topic_5_name` constitutes `r topic_5_prop` found in `r topic_5_documents` documents with FREX terms "`r topic_5_frex`" prominent in `r topic_5_countries`. The bureaucratic vocabulary (_Offic_, _Climate-Resili_, _Medium_, _Secretariat_, _Depart_) reveals institutional architecture considerations, with _Climate-Resili_ suggesting these offices focus specifically on resilience building. The _Office_ label effectively captures the administrative focus. Its presence suggests adaptation planning requires demonstrating organizational competence as much as addressing climate vulnerabilities, with countries like Bhutan and Nepal dedicating substantial discourse to institutional arrangements.

`r topic_6_name` at `r topic_6_prop` in `r topic_6_documents` documents with FREX terms "`r topic_6_frex`" dominated by `r topic_6_countries`. The terminology spans infrastructure (_Transit_), planning processes (_Task_, _Instrument_), and indigenous rights (_Indigen_, _Territori_). The _Transit_ label captures only one dimension of what appears to be a complex intersection of infrastructure development and indigenous territorial governance. Brazil and Argentina's dominance suggests this topic reflects Latin American contexts where infrastructure projects must navigate indigenous rights—a reminder that adaptation occurs within existing political geographies.

`r topic_7_name` comprises `r topic_7_prop` across `r topic_7_documents` documents with FREX terms "`r topic_7_frex`" appearing in `r topic_7_countries`. The technical climate science vocabulary (_Rcp_ for Representative Concentration Pathways, _Rainfal_, _Ensembl_, _Confid_, _Trend_) forms the clearest thematic cluster, representing the modeling and projection work underlying vulnerability assessments. The _RCP_ designation accurately identifies this as the technical scientific foundation of adaptation planning. Its limited presence and concentration in specific countries like West Bank and Gaza suggests that despite rhetoric about science-based planning, technical climate analysis remains a minor and geographically specific component of adaptation discourse.

```{r}
#| label: tbl-topics
#| tbl-cap: "Summary of adaptation discourse topics identified through structural topic modeling"
#| tbl-colwidths: [15, 10, 10, 30, 25]
#| echo: false

# Load required packages
library(knitr)
library(kableExtra)

# Create topics data frame using a loop
topics_data <- data.frame()

for(i in 1:k_topics) {
  topics_data <- rbind(topics_data, data.frame(
    Topic = get(paste0("topic_", i, "_name")),
    Proportion = get(paste0("topic_", i, "_prop")),
    Documents = get(paste0("topic_", i, "_documents")),
    `Key Terms` = get(paste0("topic_", i, "_frex")),
    Countries = get(paste0("topic_", i, "_countries"))
  ))
}

# Create formatted table (already in topic number order from loop)
kable(topics_data, 
      row.names = FALSE,
      col.names = c("Topic", "Prop", "Docs", "Terms", "Countries"))
```

## Groups

>This section examines how category membership shapes adaptation discourse, revealing that regional institutional networks exert stronger influence than actual climate vulnerabilities or economic capacity.

Regional groupings demonstrate the most powerful influence on adaptation discourse, with an average effect size of `r region_effect`—substantially higher than income (`r income_effect`), geography (`r geography_effect`), or time (`r time_effect`). This dominance suggests that regional institutional architectures—development banks, technical assistance programs, consultant networks—create convergence pressures that overwhelm both national circumstances and actual climate vulnerabilities.

Europe & Central Asia (`r europe_central_asia_abbrev`) exhibit the strongest regional effect at `r europe_central_asia_effect`, nearly half the total possible effect size. Countries in this region show dominance of `r europe_central_asia_dom` concentrated on `r europe_central_asia_topics`. Despite the region being very large and diverse, from EU members to Central Asian republics, these plans demonstrate remarkable homogeneity. The emphasis on mountain ecosystems alongside mainstream planning suggests either genuinely shared vulnerabilities or, more likely, adoption of standardized regional templates.

South Asia follows with an effect size of `r south_asia_effect`, the second highest among regions. With only four plans in the dataset, this region nonetheless shows dominance of `r south_asia_dom` focused on `r south_asia_topics`. The prominence of organizational and bureaucratic concerns suggests these nations have developed parallel institutional architectures, potentially through shared regional frameworks or common technical assistance sources.

Sub-Saharan Africa and Middle East, North Africa, Afghanistan & Pakistan both show effect sizes of `r sub_saharan_africa_effect`. Sub-Saharan Africa's dominance of `r sub_saharan_africa_dom` through `r sub_saharan_africa_topics` reflects the region's longer engagement with adaptation planning through NAPA frameworks. Meanwhile, MENA's lower dominance of `r middle_east_north_africa_afghanistan_pakistan_dom` but equal effect size suggests that while this region explores more diverse topics, countries within it still converge strongly around `r middle_east_north_africa_afghanistan_pakistan_topics`.

The relationship between discourse dominance and group effect sizes reveals how different categorization schemes create varying levels of constraint on national adaptation planning. @fig-dominance-effect maps these patterns across all subcategories.

```{r}
#| label: fig-dominance-effect
#| fig-title: "Dominance and effect"
#| fig-cap: "Dominance and Effect Size by Category. Each panel shows a different categorization scheme."
#| fig-height: 6
#| echo: false

# Load required packages
library(ggplot2)
library(ggrepel)
library(ggthemes)

# Load metrics data
metrics <- readRDS("../data/calculate_metrics.rds")
metrics_table <- metrics$data

# Prepare data for plotting
plot_data <- metrics_table[metrics_table$subcategory != "Overall" & !is.na(metrics_table$effect_size), ]
plot_data$category <- factor(plot_data$category, levels = c("Income", "Region", "Geography", "Time"))
plot_data$label <- ifelse(plot_data$significant, plot_data$subcategory, "")

ggplot(plot_data, aes(x = dominance, y = effect_size)) +
  geom_point(size = 2) +
  geom_text_repel(aes(label = subcategory_abbrev), 
            force = 3,
            point.padding = 0.5,
            box.padding = 0.5,
            size = 3) + 
  labs(title = "Dominance and effect",
       x = "Dominance", 
       y = "Estimated effect") +
  facet_wrap(~ category, scales = "fixed", ncol = 2) +
  
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  
  theme_economist_white(base_size = 11) + 
  theme(
    plot.title = element_text(size = 16, face = "bold", 
                             margin = margin(b = 15), 
                             hjust = 0.5),
    strip.text = element_text(size = 12, face = "bold"),  # Facet labels
    panel.spacing = unit(1, "lines"), # increases space between panels
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10))
  )
```

The scatter reveals no clear correlation between dominance and effect size. High-income countries show extreme topic concentration (`r high_income_dom`) but moderate group effects (`r high_income_effect`), while low-income countries display lower dominance (`r low_income_dom`) but the highest income-group effect (`r low_income_effect`). This suggests that discourse concentration and group conformity operate as independent dimensions of constraint.

Figure @fig-average presents the aggregate patterns, confirming regional categories as the primary organizing force in global adaptation discourse.

```{r}
#| label: fig-average
#| fig-cap: "Average dominance and effect size by category. Regional patterns show the strongest effects."
#| fig-height: 5
#| echo: false
# Load dependencies
library(tidyr)

# Filter and select
temp_data <- metrics_table[metrics_table$subcategory == "Overall" & metrics_table$category != "Global", ]
temp_data <- temp_data[, c("category", "dominance", "effect_size")]

# Reorder category by dominance
temp_data$category <- reorder(temp_data$category, temp_data$dominance)

# Pivot longer
plot_data <- pivot_longer(temp_data, 
                          cols = c(dominance, effect_size),
                          names_to = "metric", 
                          values_to = "value")

# Factor the metric
plot_data$metric <- factor(plot_data$metric, levels = c("effect_size", "dominance"))

# Create plot
ggplot(plot_data, aes(x = value, y = category, fill = metric)) +
  geom_col(position = "dodge", alpha = 0.7) +
  geom_text(aes(label = paste0(round(value * 100), "%")), 
            position = position_dodge(width = 1),
            hjust = -0.1, 
            size = 4) +
  labs(title = "Average across categories",
       x = "Average Value", 
       y = NULL) +                          
  scale_x_continuous(
    expand = expansion(mult = c(0, 0.1)),
    labels = percent_format(accuracy = 1)) +
  
  theme_economist_white(base_size = 11) + 
  scale_fill_economist(name = "", labels = c("Effect Size", "Dominance")) +
  theme(
    legend.position = c(0.98, 0.02), # Bottom-right corner
    legend.justification = c(1, 0), # Anchor to bottom-right
    legend.background = element_rect(fill = "transparent", color = NA),
    legend.title = element_blank(),
    legend.text = element_text(size = 10), # Match base_size
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    plot.title = element_text(size = 16, face = "bold", 
                             margin = margin(b = 15), 
                             hjust = 0.5)
  )
```

The consistently high values across both metrics for regions, compared to the variation in other categories, reinforces that institutional geography trumps physical geography in shaping how countries articulate climate responses.

Income categories show an overall effect size of `r income_effect`, revealing counterintuitive patterns. Low-income countries demonstrate the highest effect within this category at `r low_income_effect`—nearly double that of upper-middle income countries (`r upper_middle_income_effect`). This extreme convergence among poor countries, despite their geographic and climatic diversity, suggests powerful standardizing forces operate through development assistance and technical support mechanisms.

Low-income countries focus on `r low_income_topics` with dominance of `r low_income_dom`. The prominence of NAPA alongside mainstream planning reflects how these countries must demonstrate fluency in both historical and contemporary adaptation frameworks. Their limited autonomy in discourse construction likely reflects dependence on external technical assistance and the need to align with donor priorities.

High-income countries show dominance of `r high_income_dom` concentrated on `r high_income_topics`, yet their effect size of `r high_income_effect` sits mid-range. This combination of extreme topic concentration with moderate group coherence suggests that while wealthy countries converge on certain topics, they maintain some flexibility in how they articulate them.

Upper-middle income countries exhibit the most autonomy, with the lowest dominance (`r upper_middle_income_dom`) and smallest effect size (`r upper_middle_income_effect`) focused on `r upper_middle_income_topics`. These countries appear to occupy a sweet spot—developed enough to resist full standardization but not so wealthy as to converge on post-material concerns.

Geographic vulnerability classifications show an overall effect of `r geography_effect`, surprisingly lower than institutional categories. Countries sharing similar climate vulnerabilities show less discourse similarity than countries in the same World Bank region, challenging assumptions about environmental determinism in adaptation planning.

Small Island Developing States exhibit the highest dominance in the entire dataset at `r sids_dom`, focusing on `r sids_topics` with an effect size of `r sids_effect`. This extreme concentration reflects genuinely shared existential threats from sea-level rise and tropical cyclones. Yet even here, mainstream planning dominates alongside obviously relevant island concerns, suggesting institutional requirements constrain even the most vulnerable nations.

Landlocked Developing Countries show moderate patterns with dominance of `r lldc_dom` through `r lldc_topics` and effect size of `r lldc_effect`. Surprisingly, neither mountain nor water management topics feature prominently despite these being defining challenges. Instead, organizational and mainstream topics dominate, indicating institutional templates override geographic specificities.

Other countries display the lowest geographic dominance at `r other_dom` with an effect size of `r other_effect` focused on `r other_topics`. This baseline group's lower convergence might reflect greater heterogeneity when countries lack strong categorical identity, though procedural topics still predominate.

Temporal categories show the weakest effects at `r time_effect`, suggesting adaptation discourse has not substantially evolved over the decade of NAP development. While individual periods show variation—middle period plans (2019-2021) demonstrating highest dominance at `r middle_dom` focused on `r middle_topics`—the overall temporal effect remains minimal.

Early submissions (2015-2018) show dominance of `r early_dom` with the lowest temporal effect of `r early_effect` on `r early_topics`. Middle period submissions exhibit the highest temporal convergence at `r middle_effect`, coinciding with clearer guidelines and more rigid finance requirements. Late submissions (2022-2025) display the lowest dominance at `r late_dom` but maintain substantial effect size of `r late_effect` on `r late_topics`.

The temporal patterns reveal no clear trajectory toward either convergence or divergence, suggesting that once established, adaptation planning frameworks persist regardless of accumulating experience or evolving climate science. 

```{r}
#| label: tbl-groups
#| tbl-cap: "Discourse patterns by category membership showing dominance and effect sizes"
#| tbl-colwidths: [15, 30, 15, 10, 30]
#| echo: false

# Load required packages
library(knitr)
library(kableExtra)

# Define all subcategories and their parent categories
subcategories <- list(
  Income = c("high_income", "upper_middle_income", "lower_middle_income", "low_income"),
  Region = c("europe_central_asia", "latin_america_caribbean", "south_asia", 
             "sub_saharan_africa", "east_asia_pacific", 
             "middle_east_north_africa_afghanistan_pakistan"),
  Geography = c("sids", "lldc", "other"),
  Time = c("early", "middle", "late")
)

# Create clean name mapping
clean_names <- c(
  "high_income" = "High Income",
  "upper_middle_income" = "Upper-Middle Income",
  "lower_middle_income" = "Lower-Middle Income", 
  "low_income" = "Low Income",
  "europe_central_asia" = "Europe & Central Asia",
  "latin_america_caribbean" = "Latin America & Caribbean",
  "south_asia" = "South Asia",
  "sub_saharan_africa" = "Sub-Saharan Africa",
  "east_asia_pacific" = "East Asia & Pacific",
  "middle_east_north_africa_afghanistan_pakistan" = "MENA, Afghanistan & Pakistan",
  "sids" = "Small Island Developing States",
  "lldc" = "Landlocked Developing Countries",
  "other" = "Other",
  "early" = "Early (2015-2018)",
  "middle" = "Middle (2019-2021)",
  "late" = "Late (2022-2025)"
)

# Create data frame with single loop
group_data <- data.frame()

for(cat in names(subcategories)) {
  for(subcat in subcategories[[cat]]) {
    group_data <- rbind(group_data, data.frame(
      Category = cat,
      Subcategory = clean_names[subcat],
      Dominance = get(paste0(subcat, "_dom")),
      `Effect Size` = get(paste0(subcat, "_effect")),
      `Top Topics` = get(paste0(subcat, "_topics"))
    ))
  }
}

# Create formatted table
kable(group_data, 
      format = "markdown",
      row.names = FALSE,
      escape = FALSE,
      col.names = c("Category", "Group", "Dominance", "Effect", "Topics"))
```