```{r load data, include=FALSE}
variables <- readRDS("../data/load_variables.rds")
vars <- variables$data
list2env(vars, envir = .GlobalEnv)
```

# Findings {#sec-findings}

>This chapter presents the findings from the STM. It finds that the most important topic is `r topic_8_name`, and that there is a high concentration across the board, with the highest estimated effect coming from membership in regional groups (`r region_effect`). 

The findings show a high level of concentration across the board, suggesting that the NAPs are concentrated somehow.

In the first section I go through the topic findings, their proportion of the whole corpus and the number of effective documents they are in. I find that while `r topic_8_name` is `r topic_8_prop` of the corpus, it is also present to a meaningful degree in `r topic_8_documents` of the `r n_documents_after_aggregation` in the corpus. 

In the second section I go through the group findings. Here, surprisingly or not, regional factors (`r region_effect`) have a larger estimated effect than income (`r income_effect`) or geography (`r geography_effect`). 

## Topics


>

The remaining topics address diverse vulnerabilities and planning dimensions, though all filtered through institutional lenses and operating at much smaller scales. `r topic_1_name`, representing `r topic_1_prop` of the corpus, exhibits FREX terms "`r topic_1_frex`" with highest representation in `r topic_1_countries`. Despite its naming after the NAPA (National Adaptation Programme of Action) framework, closer examination reveals this topic encompasses vulnerability concerns through terms including "poverti" (poverty), "poor" (economic status), "children", "women", "migrat" (population movement), and "sanit" (sanitation). This reflects how adaptation in economically constrained contexts becomes framed primarily as poverty reduction rather than environmental response.

`r topic_2_name`, comprising `r topic_2_prop` of the corpus, exhibits FREX terms "`r topic_2_frex`" with highest representation in `r topic_2_countries`. This topic captures marine and coastal environments through "sea" (ocean systems), "coastal" (shoreline areas), "rise" (sea-level rise), "cyclon" (tropical cyclones), and "slr" (sea-level rise abbreviation). The geographic specificity to small island and coastal nations validates how certain environmental challenges generate recognizable thematic clusters, though still expressed through standardized vulnerability assessment frameworks.

`r topic_3_name`, the least prevalent at `r topic_3_prop`, shows FREX terms "`r topic_3_frex`" with clustering around `r topic_3_countries`. This topic encompasses water infrastructure and management systems through "irrig" (irrigation systems), "hydrometeorolog" (meteorological systems), "mountain" (highland environments), "wastewat" (wastewater management), and "basin" (watershed areas). The country pattern suggests technical water management approaches that transcend developmental contexts, indicating how engineering frameworks standardize responses to hydrological vulnerabilities.

`r topic_4_name` at `r topic_4_prop` shows FREX terms "`r topic_4_frex`" with clustering around `r topic_4_countries`. This topic encompasses rural livelihoods and agricultural systems through terms including "livestock" (animal husbandry), "farm" (agricultural systems), "harvest" (crop production), "pest" (agricultural threats), and "watersh" (watershed management). The countries represented—nations with large rural populations and agriculture-dependent economies—confirm how economic structures influence which climate vulnerabilities receive sustained attention in planning documents.

`r topic_5_name`, comprising `r topic_5_prop` of the corpus, complements the procedural focus through organizational and bureaucratic dimensions. The FREX terms "`r topic_5_frex`" with clustering around `r topic_5_countries` emphasize institutional arrangements through terms like "offic" (offices), "organis" (organizations), "depart" (departments), "output" (deliverables), and "partnership" (collaborative frameworks). This topic captures the administrative infrastructure deemed necessary for adaptation implementation, suggesting that organizational capacity-building has become as important as addressing climate impacts themselves.

`r topic_6_name` at `r topic_6_prop` presents an interesting anomaly. FREX terms "`r topic_6_frex`" with highest representation in `r topic_6_countries` initially suggest transition planning, but closer examination reveals infrastructure development concerns. The focus on "transit" and "line" as urban infrastructure terms, combined with "executiv" (executive) and "instrument" (policy instruments), alongside "indigen" (indigenous) and "territori" (territorial), reflects Brazil's specific context where indigenous populations represent significant stakeholders in infrastructure development. This misalignment between automated naming and actual content highlights how regional specificities can be obscured within standardized topic frameworks.

`r topic_7_name`, at `r topic_7_prop` of the corpus, presents the technical modeling dimension of adaptation planning. FREX terms "`r topic_7_frex`" encompass climate science vocabulary including "rcp" (Representative Concentration Pathways), "ensembl" (ensemble modeling), "centuri" (century timeframes), "precipit" (precipitation), and "confid" (confidence intervals). The surprising country clustering around `r topic_7_countries` reveals how technical assessment frameworks operate across diverse political contexts. Manual review of the Palestinian NAP confirms considerable focus on precipitation modeling while notably excluding any mention of Israel, perhaps indicating how technical language can obscure political realities.

`r topic_8_name`, representing `r topic_8_prop` of the corpus, dominates adaptation discourse to an extraordinary degree—more than three times larger than any other topic. Its FREX terms "`r topic_8_frex`" suggest that this theme centers on the planning of planning itself. Here, knowledge transfer (*learn*) plays a central role, as well as the international structures for climate change, the Paris agreement (*pari* and *agreement*). The only term up for debate is progress, which can hint at progress towards climate related global goals, or progress in a more general sense. That the main countries, `topic_8_countries` are both in Europe & Central Asia, the most concentrated region, as we will see below, although their dominance values are not the highest we will observe in the data. 

```{r}
#| label: fig-topics
#| fig-cap: "Topics by proportion and document coverage. The mainstream topic dominates both dimensions."
#| fig-height: 5
#| echo: false

# Load dependencies
library(ggplot2)
library(ggrepel)
library(ggthemes)

# Load data
topics <- readRDS("../data/name_topics.rds")
topics_table <- topics$data

# Create plot
ggplot(topics_table, aes(x = topic_proportion, y = effective_documents)) +
  geom_point(size = 2) +
  geom_text_repel(aes(label = topic_name), 
            force = 4,
            point.padding = 0.5,
            box.padding = 0.5,
            size = 5) +
  labs(title = "Topic distrobution",
       x = "Topic Proportion",
       y = "Effective Documents"
  ) +
  theme_economist_white(base_size = 10) + 
  theme(legend.position = "none",
        plot.title = element_text(size = 16, face = "bold", 
                             margin = margin(b = 15), 
                             hjust = 0.5),
        axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 10)))
```

## Groups


The relationship between national wealth and adaptation discourse reveals unexpected patterns. High-income countries show the most extreme dominance at `r high_income_dom`, with their discourse overwhelmingly concentrated on `r high_income_topics`. This represents nearly two-thirds of their total discourse focused on just three topics—a remarkable concentration for countries presumably capable of addressing multiple dimensions simultaneously. Their effect size of `r high_income_effect` sits in the middle range, neither the highest nor lowest among income groups.

Upper-middle income countries present an interesting contrast, exhibiting the lowest dominance among income categories at `r upper_middle_income_dom` with emphasis on `r upper_middle_income_topics`. Notably, they also show the smallest effect size at `r upper_middle_income_effect`, suggesting these countries maintain the most varied approaches within their group. This combination of lower dominance and lower effect size hints at greater flexibility in how these countries structure their adaptation discourse.

```{r}
#| label: fig-dominance-effect
#| fig-title: "Dominance and effect"
#| fig-cap: "Dominance and Effect Size by Category. Each panel shows a different categorization scheme."
#| fig-height: 6
#| echo: false
# Load required packages
library(ggplot2)
library(ggrepel)
library(ggthemes)
# Load metrics data
metrics <- readRDS("../data/calculate_metrics.rds")
metrics_table <- metrics$data
# Prepare data for plotting
plot_data <- metrics_table[metrics_table$subcategory != "Overall" & !is.na(metrics_table$effect_size), ]
plot_data$category <- factor(plot_data$category, levels = c("Income", "Region", "Geography", "Time"))
plot_data$label <- ifelse(plot_data$significant, plot_data$subcategory, "")

ggplot(plot_data, aes(x = dominance, y = effect_size)) +
  geom_point(size = 2) +
  geom_text_repel(aes(label = subcategory_abbrev), 
            force = 3,
            point.padding = 0.5,
            box.padding = 0.5,
            size = 3) + 
  labs(title = "Dominance and effect",
       x = "Dominance", 
       y = "Estimated effect") +
  facet_wrap(~ category, scales = "fixed", ncol = 2) +
  
  theme_economist_white(base_size = 12) + 
  theme(
    plot.title = element_text(size = 16, face = "bold", 
                             margin = margin(b = 15), 
                             hjust = 0.5),
    strip.text = element_text(size = 12, face = "bold"),  # Facet labels
    panel.spacing = unit(1, "lines"), # increases space between panels
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10))
  )
```

Lower-middle income countries display moderate patterns, with dominance of `r lower_middle_income_dom` through `r lower_middle_income_topics`. Their effect size of `r lower_middle_income_effect` falls between the extremes, suggesting moderate group coherence without excessive rigidity.

Low-income countries exhibit dominance of `r low_income_dom` with distinctively different top topics: `r low_income_topics`. Significantly, their effect size of `r low_income_effect` stands out as the highest among all income categories—nearly double that of upper-middle income countries. This extreme effect size suggests low-income countries show remarkably similar discourse patterns to each other, with limited variation between individual countries in this group.

The non-linear relationship between income and these metrics challenges simple explanations. While one might expect a clear gradient from rich to poor, the actual pattern shows upper-middle income countries with the most autonomy (lowest effect size) while low-income countries show the strongest group constraints (highest effect size). The overall income category shows dominance of `r income_dom` with an average effect size of `r income_effect`, confirming that wealth levels significantly shape adaptation discourse, though not in straightforward ways.

Regional groupings reveal the strongest patterns in the entire dataset, with an overall dominance of `r region_dom` and average effect size of `r region_effect`—the highest among all category types. This suggests regional institutional networks and shared frameworks powerfully shape how countries conceptualize adaptation, potentially more than actual climate vulnerabilities.

Europe & Central Asia demonstrates extraordinary convergence, showing dominance of `r europe_central_asia_dom` through `r europe_central_asia_topics`. The effect size of `r europe_central_asia_effect` ranks as the highest regional effect and among the highest in the entire dataset. This exceptional convergence occurs despite the region encompassing diverse contexts from EU members to Central Asian republics, suggesting powerful standardizing forces at work.

Latin America & Caribbean exhibits similarly high dominance at `r latin_america_caribbean_dom` with emphasis on `r latin_america_caribbean_topics`. The prominence of transit alongside mainstream topics distinguishes this region's discourse. With an effect size of `r latin_america_caribbean_effect`, the region shows significant but not extreme group convergence, allowing some space for national variation within shared regional frameworks.

South Asia, despite containing only four plans, displays remarkable uniformity with dominance of `r south_asia_dom` focused on `r south_asia_topics`. The high effect size of `r south_asia_effect` indicates strong regional coherence, though the small sample size warrants cautious interpretation. The emphasis on office and organizational topics alongside mainstream planning suggests shared institutional concerns across the region.

Sub-Saharan Africa, the most geographically and climatically diverse region, nonetheless shows significant convergence with dominance of `r sub_saharan_africa_dom` through `r sub_saharan_africa_topics`. The effect size of `r sub_saharan_africa_effect` reveals substantial regional patterns despite enormous variation in actual climate challenges—from Sahelian droughts to tropical flooding to highland agriculture. The prominence of NAPA topics alongside mainstream planning reflects the region's longer engagement with adaptation frameworks.

East Asia & Pacific exhibits dominance of `r east_asia_pacific_dom` through `r east_asia_pacific_topics`, combining procedural concerns with obvious regional vulnerabilities. The effect size of `r east_asia_pacific_effect` represents the lowest regional effect, suggesting more diversity within this group—perhaps reflecting the vast differences between Pacific island states and continental Asian countries included in this category.

Middle East, North Africa, Afghanistan & Pakistan shows the lowest regional dominance at `r middle_east_north_africa_afghanistan_pakistan_dom` with emphasis on `r middle_east_north_africa_afghanistan_pakistan_topics`. Despite this lower dominance, the effect size of `r middle_east_north_africa_afghanistan_pakistan_effect` remains substantial, indicating that even in this politically and environmentally diverse region, countries converge around similar adaptation frameworks. The prominence of RCP (climate projections) as the lead topic distinguishes this region's technical emphasis.

All regional subcategories demonstrate statistical significance (p < 0.05), with no exceptions. This universal significance, combined with consistently high effect sizes, suggests regional institutional architectures—development banks, technical assistance programs, consultant networks—create powerful convergence pressures that shape adaptation discourse regardless of actual environmental differences.

Geographic vulnerability classifications reveal how shared environmental exposures translate into discourse patterns, with overall dominance of `r geography_dom` and average effect size of `r geography_effect`. These categories, based on recognized climate vulnerabilities, might be expected to show the strongest convergence, yet they demonstrate more modest patterns than regional groupings.

Small Island Developing States exhibit the highest dominance in the entire dataset at `r sids_dom`, focusing overwhelmingly on `r sids_topics`. This extreme concentration—over two-thirds of discourse in just three topics—reflects how acute, existential climate threats create highly focused adaptation narratives. The effect size of `r sids_effect` confirms strong group coherence, as these nations face genuinely similar challenges from sea-level rise and tropical cyclones. Yet even here, the mainstream planning topic dominates alongside obviously relevant island and cyclone concerns.

Landlocked Developing Countries show moderate dominance of `r lldc_dom` through emphasis on `r lldc_topics`. The effect size of `r lldc_effect` indicates modest group patterns. Surprisingly, neither mountain nor water management topics feature prominently despite these being defining challenges for landlocked nations. Instead, organizational and procedural topics dominate, suggesting institutional concerns overshadow geographic specificities.

Other countries—those without special geographic designations—display the lowest geographic dominance at `r other_dom` with top topics of `r other_topics`. The effect size of `r other_effect` represents the weakest geographic pattern, though still statistically significant. This baseline group's lower convergence might reflect greater heterogeneity when countries lack a strong categorical identity, yet procedural topics still feature prominently even in this diverse group.

The pattern where SIDS show extreme dominance while other categories show moderate levels suggests that only the most acute, internationally recognized vulnerabilities generate strongly concentrated discourse. Countries with less visible or more complex vulnerabilities default to institutional frameworks rather than geography-specific approaches.

Analysis of submission timing reveals how adaptation discourse has evolved over the decade-long development of NAPs, with overall temporal dominance of `r time_dom` and average effect size of `r time_effect`—the lowest among all category types, suggesting time matters less than income, region, or geography in shaping discourse.

Early submissions (2015-2018) show moderate dominance of `r early_dom` with `r early_topics`. The effect size of `r early_effect`—the lowest temporal effect—suggests early adopters maintained relatively diverse approaches, perhaps because frameworks were still emerging and countries had more freedom to interpret adaptation planning according to national preferences.

Middle period submissions (2019-2021) exhibit the highest temporal dominance at `r middle_dom` through focus on `r middle_topics`. More significantly, the effect size of `r middle_effect` represents the strongest temporal convergence, suggesting this period saw the most standardization. This coincides with the establishment of clearer guidelines, increased technical assistance, and more rigid requirements for accessing climate finance. Countries submitting during this period show remarkably similar discourse patterns.

Late submissions (2022-2025) present an interesting shift, displaying the lowest dominance at `r late_dom` with `r late_topics`. This suggests some diversification in recent plans. However, the effect size of `r late_effect` remains substantial—nearly as high as the middle period—indicating that while recent plans explore slightly more varied topics, they still operate within strongly constraining frameworks. The continued prominence of mainstream planning across all periods shows that procedural requirements have become institutionalized rather than evolving.

The temporal progression reveals no clear trend toward either convergence or divergence. Instead, it suggests waves of standardization, with the middle period representing peak conformity while recent submissions show modest diversification within persistent structural constraints.

```{r}
#| label: fig-average
#| fig-cap: "Average dominance and effect size by category. Regional patterns show the strongest effects."
#| fig-height: 5
#| echo: false
# Load dependencies
library(tidyr)

# Filter and select
temp_data <- metrics_table[metrics_table$subcategory == "Overall" & metrics_table$category != "Global", ]
temp_data <- temp_data[, c("category", "dominance", "effect_size")]

# Reorder category by dominance
temp_data$category <- reorder(temp_data$category, temp_data$dominance)

# Pivot longer
plot_data <- pivot_longer(temp_data, 
                          cols = c(dominance, effect_size),
                          names_to = "metric", 
                          values_to = "value")

# Factor the metric
plot_data$metric <- factor(plot_data$metric, levels = c("effect_size", "dominance"))

# Create plot
ggplot(plot_data, aes(x = value, y = category, fill = metric)) +
  geom_col(position = "dodge", alpha = 0.7) +
  geom_text(aes(label = round(value, 2)), 
            position = position_dodge(width = 1),
            hjust = -0.1, 
            size = 4) +
  labs(title = "Average across categories",
       x = "Average Value", 
       y = NULL) +                          
  scale_x_continuous(expand = expansion(mult = c(0, 0.1))) +
  
  theme_economist_white(base_size = 10) + 
  scale_fill_economist(name = "", labels = c("Effect Size", "Dominance")) +
  theme(
    legend.position = c(0.98, 0.02),                    # Bottom-right corner
    legend.justification = c(1, 0),                     # Anchor to bottom-right
    legend.background = element_rect(fill = "transparent", color = NA),
    legend.title = element_blank(),
    legend.text = element_text(size = 10),              # Match base_size
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    plot.title = element_text(size = 16, face = "bold", 
                             margin = margin(b = 15), 
                             hjust = 0.5)
  )
```
